---
title: "STAT 600 - HW 3"
author: "Kevin Korsurat"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(knitr)
library(Rcpp)
library(RcppArmadillo)
library(foreach)
library(doParallel)
library(ggplot2)
library(latex2exp)


path <- "/Users/kevin-imac/Desktop/Github - Repo/"
if(! file.exists(path)){
  path <- "/Users/kevinkvp/Desktop/Github Repo/"
}

# sourceCpp(paste0(path, "HW3EM/src/main.cpp"))
```

All Rcpp/RcppArmadillo can be found in my [\textcolor{red}{GitHub}](https://github.com/skorsu/HW3EM).

```{r}

### User-defined functions -----------------------------------------------------
meanSD <- function(x, dplace = 5){
  mm <- round(mean(x), digits = dplace)
  ss <- round(sd(x), digits = dplace)
  paste0(mm, " (SD = ", ss, ")")
}
```

## Derivation

First, I will derive the log-likelihood function given the completed data, $l\left(p, \lambda, \mu | \boldsymbol{Y}, \boldsymbol{\delta} \right)$.

$$\begin{aligned}
l\left(p, \lambda, \mu | \boldsymbol{Y}, \boldsymbol{\delta} \right) &= \log\left(L\left(p, \lambda, \mu | \boldsymbol{Y}, \boldsymbol{\delta} \right)\right) \\
&= \log\left(\prod_{i=1}^{n} p\left(Y_{i}, \delta_{i}| p, \lambda, \mu\right)\right) \\
&= \log\left(\prod_{i=1}^{n} \left(p\lambda\exp\left(-\lambda y_{i}\right)\right)^{\delta_{i}} \left(\left(1-p\right)\mu\exp\left(-\mu y_{i}\right)\right)^{1-\delta_{i}} \right) \\
&= \sum_{i=1}^{n}\log\left\{\left(p\lambda\exp\left(-\lambda y_{i}\right)\right)^{\delta_{i}} \left(\left(1-p\right)\mu\exp\left(-\mu y_{i}\right)\right)^{1-\delta_{i}}\right\} \\
&= \sum_{i=1}^{n}\left[\delta_{i}\left\{-\lambda y_{i} + \log\left(p \lambda\right)\right\} + \left(1 - \delta_{i}\right)\left\{-\mu y_{i} + \log\left(\left(1 - p\right)\mu\right)\right\}\right]
\end{aligned}$$

Then, the E-step in the EM algorithm is shown below.

$$\begin{aligned}
Q\left(p, \lambda, \mu | p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right) &= E\left[l\left(p, \lambda, \mu | \boldsymbol{Y}, \boldsymbol{\delta} \right) | \boldsymbol{Y}, p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right] \\
&= E\left[\sum_{i=1}^{n}\left[\delta_{i}\left\{-\lambda y_{i} + \log\left(p \lambda\right)\right\} + \left(1 - \delta_{i}\right)\left\{-\mu y_{i} + \log\left(\left(1 - p\right)\mu\right)\right\}\right] \Bigg | \boldsymbol{Y}, p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right] \\
&= \sum_{i=1}^{n} E\left[\delta_{i}\left\{-\lambda y_{i} + \log\left(p \lambda\right)\right\} + \left(1 - \delta_{i}\right)\left\{-\mu y_{i} + \log\left(\left(1 - p\right)\mu\right)\right\} \Bigg | y_{i}, p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right] \\
&= \sum_{i=1}^{n} \left\{\hat\delta^{(t)}_{i}\left[-\lambda y_{i} + \log\left(p \lambda\right)\right]  + \left(1 - \hat\delta^{(t)}_{i}\right)\left[-\mu y_{i} + \log\left(\left(1 - p\right)\mu\right)\right]\right\}
\end{aligned}$$

where $\hat\delta^{(t)}_{i} = E\left[\delta_{i}\right| y_{i}, p^{(t)}, \lambda^{(t)}, \mu^{(t)}] = \frac{p^{(t)}\lambda^{(t)} \exp{(-\lambda^{(t)} y_{i})}}{p^{(t)}\lambda^{(t)} \exp{(-\lambda^{(t)} y_{i})}+(1-p^{(t)})\mu^{(t)}\exp{(-\mu^{(t)} y_{i})}}$

Then, I will derive the M-step by starting the finding the $p^{(t+1)}$.

$$\begin{aligned}
\frac{d}{dp}Q\left(p, \lambda, \mu | p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right) &\stackrel{\mbox{set}}{=} 0 \\
\sum_{i=1}^{n}\left[\frac{\hat\delta_{i}^{(t)}}{p} - \frac{1 - \hat\delta_{i}^{(t)}}{1 - p}\right] &= 0 \\
\sum_{i=1}^{n}\left[(1-p)\hat\delta_{i}^{(t)} - \left(1 - \hat\delta_{i}^{(t)}\right)p\right] &= 0 \\
\sum_{i=1}^{n}\hat\delta_{i}^{(t)} -np &= 0 \\
p^{(t+1)} &= \frac{1}{n}\sum_{i=1}^{n}\hat\delta_{i}^{(t)}
\end{aligned}$$

Below is the derivation for $\lambda^{(t+1)}$.

$$\begin{aligned}
\frac{d}{d\lambda}Q\left(p, \lambda, \mu | p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right) &\stackrel{\mbox{set}}{=} 0 \\
\sum_{i=1}^{n}\left[-\hat\delta_{i}^{(t)}y_{i} + \frac{\hat\delta_{i}^{(t)}}{\lambda}\right] &= 0 \\
-\lambda \sum_{i=1}^{n} \hat\delta_{i}^{(t)} y_{i} + \sum_{i=1}^{n}\hat\delta_{i}^{(t)} &= 0 \\
\lambda^{(t+1)} &= \frac{\sum_{i=1}^{n}\hat\delta_{i}^{(t)}}{\sum_{i=1}^{n}\hat\delta_{i}^{(t)}y_{i}}
\end{aligned}$$

Lastly, this is the derivation for $\mu^{(t+1)}$

$$\begin{aligned}
\frac{d}{d\mu}Q\left(p, \lambda, \mu | p^{(t)}, \lambda^{(t)}, \mu^{(t)} \right) &\stackrel{\mbox{set}}{=} 0 \\
\sum_{i=1}^{n}\left[\left(1-\hat\delta_{i}^{(t)}\right)y_{i} + \frac{1-\hat\delta_{i}^{(t)}}{\mu}\right] &= 0 \\
-\mu \sum_{i=1}^{n} \left(1 - \hat\delta_{i}^{(t)}\right) y_{i} + \sum_{i=1}^{n}\left(1-\hat\delta_{i}^{(t)}\right) &= 0 \\
\mu^{(t+1)} &= \frac{\sum_{i=1}^{n}\left(1-\hat\delta_{i}^{(t)}\right)}{\sum_{i=1}^{n}\left(1-\hat\delta_{i}^{(t)}\right)y_{i}}
\end{aligned}$$

\newpage

## Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE} 

```
